---
title: "Using a custom statistical test"
author: Jean Monlong
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
  github_document: default
vignette: >
  %\VignetteIndexEntry{Using a custom statistical test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(StoatPlot)
library(dplyr)
library(knitr) ## for table outputs
library(ggplot2)
library(car) ## to test for specific coeff in regression
```

We use the small test data from the package.

```{r}
test_data_dir <- system.file('extdata', package='StoatPlot')
gt_fn = file.path(test_data_dir, 'snarl_genotypes.sorted.tsv.gz')
pheno_fn = file.path(test_data_dir, 'phenotype.quantitative.tsv')
```

## Read the phenotype file

```{r}
pheno = readr::read_tsv(pheno_fn)

pheno |> head() |> kable()
```

## Make two random covariables

```{r}
covars = tibble(SAMPLE=pheno$SAMPLE,
                sex=sample(0:1, nrow(pheno), replace=TRUE),
                PC1=rnorm(nrow(pheno)))
```

## Extract snarls in a region

The genotype matrix can get very big and we don't necessarily want to load it entirely in memory. 
Instead, we can load the snarl information in chunks.

Let's extract the snarls overlapping `ref:20000-50000`.

```{r}
chunk.l = import_genotype_chunk(gt_fn, 'ref', 20000, 50000)
length(chunk.l)
```

## Building a test table for one snarl

Let's take the first snarl info and make a test table by combining with the phenotype.

```{r}
df = combine_chunk_phenotype_covars(chunk.l[[1]]$GT, pheno, covars)

df |> head() |> kable()
```

This table could be used for a test, for example fitting a regression model.

In STOAT, we also preprocess it a bit to remove constant variables, obvious colinearity.
The *sample* column is also not used in the regression.
We've created a function to mimick the table preparation used by STOAT.

```{r}
df = preprocess_test_table(df)

df |> head() |> kable()
```

Now the table only contains:

- a `phenotype` variable
- one or several *allele* variables prefixed by `al`
- one or several *covariables* prefixed by `covar`
- potentially, a new covariable `total_allele_count` to control for the effect of the parent snarl (to be used like a covariable). See example below.

In STOAT, we fit a model and test if any of the coefficients for the *allele* variables are different from 0.

## Example of a nested snarl

If a snarl is nested, a new covariable names *total_allele_count* is added by `preprocess_test_table` to control for the effect of the parent snarl.
This new variables should be used like a covariable.

```{r}
df = combine_chunk_phenotype_covars(chunk.l[[34]]$GT, pheno, covars)
df = preprocess_test_table(df)
df |> head() |> kable()
```

## Compute p-values using linear regression

For example, let's compute a pvalue by fitting a linear regression model for each snarl.

```{r}
pv.df = lapply(chunk.l, function(snarl_info) {
  ## combine allele count with the phenotype
  df = combine_chunk_phenotype_covars(snarl_info$GT, pheno, covars)
  df = preprocess_test_table(df)

  ## remove the sample name column
  df$sample = NULL
  
  ## run the regression
  model = lm(phenotype ~ ., data=df)
  ## null hypothesis is all allele coeffs are 0
  null.h = paste0(grep('^al', colnames(df), value=TRUE), '=0')
  lh.o = linearHypothesis(model, null.h)
  pv = lh.o[['Pr(>F)']][2]

  ## return a data.frame with the snarl information and the test result
  return(tibble(START_NODE=snarl_info$START_NODE, END_NODE=snarl_info$END_NODE,
                P=pv))
}) |> bind_rows()

pv.df |> head() |> kable()
```
